MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-03-13 MAESTROeX_I/equations.tex
more updates

\section{Governing Equations}
The derivation of the low Mach number hydrodynamics equations is given in Papers I-III.
We take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field.  This is done by taking
the Langrangian derivative of the equation of state for pressure as a function of the
density, composition, and enthalpy, enforcing the constraint that the pressure is
a prescribed function of altitutde and time based on the hydrostatic equilibrium condition,
and substituting the equations for species and enthalpy in the resulting Lagrangian derivative.
The resulting equations are
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc,\label{eq:enthalpy}
\end{eqnarray}
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
Here $\Hnuc$ is the nuclear energy generation rate per unit mass.
The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, 
which is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by the equation of state which we
express as a divergence constraint on the velocity field (\ref{eq:U divergence}),
\MarginPar{This is different from the Jacobs 2016 paper where we switched to alt\_energy\_fix.}
where $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

>>>>

\section{Governing Equations}
Low Mach number models for reacting flow were originally derived using asymptotic analysis
\citep{rehm1978equations,majda1985derivation} and used in terrestrial combustion applications
\citep{knio1999semi,day2000numerical}.  These models have been extended to nuclear flames
in astrophysical environments using adaptive algorithms in space and time \citep{Bell:2004}.
In Papers I-III, we derived a model and algorithm suitable for stratified astrophysical flow.
In our model, we take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field.
The resulting model is a series of evolution equations for mass, momentum, and energy, subject
to this constraint:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc,\label{eq:enthalpy}
\end{eqnarray}
\begin{equation}
\nabla\cdot(\beta_0\Ub) = \beta_0\left(S - \frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial t}\right).\label{eq:U divergence}
\end{equation}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$.  The species are constrained
such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
Here $\Hnuc$ is the nuclear energy generation rate per unit mass.
The pressure is decomposed into a hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$|\pi|/p_0 = \mathcal{O}(M^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  We also define a base state density, $\rho_0 = \rho_0(r,t)$, 
which is in hydrostatic equilibrium with $p_0$, i.e., 
$\nabla p_0 = -\rho_0 g\eb_r$, where $g=g(r,t)$ is
the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the
outward radial direction.

Mathematically, equations (\ref{eq:momentum})-(\ref{eq:enthalpy}) must still be closed by an equation of state.  
This is done by taking the Langrangian derivative of the equation of state for pressure as a function of the density, composition, and enthalpy, requiring that the pressure is a prescribed function of altitutde and time based on the hydrostatic equilibrium condition.
Details are given in Papers I and II.
After substuting the equations for mass and energy into the Lagrangian derivative, the result
is a divergence constraint on the velocity field (\ref{eq:U divergence}),
where $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho)$ at constant entropy.
The expansion term, $S$, incorporates local compressibility effects due to heat release from reactions, compositional changes, and external sources,
\begin{equation}
S = -\sigma\sum_k\xi_k\omegadot_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\omegadot_k + \sigma\Hnuc,\label{eq:S}
\end{equation}
where $p_{X_k} \equiv \left. \partial p / \partial X_k
\right|_{\rho,T,X_{j,j\ne k}}$, $\xi_k \equiv \left. \partial h /
\partial X_k \right |_{p,T,X_{j,j\ne k}},
p_\rho \equiv \left.
\partial p/\partial \rho \right |_{T, X_k}$, and $\sigma \equiv
p_T/(\rho c_p p_\rho)$, with $p_T \equiv \left. \partial p / \partial
T \right|_{\rho, X_k}$ and $c_p \equiv \left.  \partial h / \partial T
\right|_{p,X_k}$ is the specific heat at constant pressure.

### 2019-03-13 MAESTROeX_I/intro.tex
more updates

Many astrophysical flows are highly subsonic; sound waves carry sufficiently little energy that they do not significantly affect the convective dynamics of the system.
In many of these flows, modeling long-time convective dynamics are of interest, and numerical approaches based on compressible hydrodynamics are intractable, even on modern supercomputers.
One approach to this problem is to use low Mach number models.
In a low Mach number approach, sound waves are eliminated from the governing equations while retaining compressibilitiy effects due to, e.g., nuclear energy release, compositional changes, and thermal diffusion.
The resulting model can be numerically integrated with much larger time steps than a compressible model.

Previously, we have developed the low Mach number astrophysical solver, MAESTRO.
Low Mach number models have been developed for a variety of contexts including combustion \citep{day2000numerical}, atmospherics \citep{duarte2015low}, and elastic solids \citep{abbate2017all}.
The low Mach number model in MAESTRO is unique in that it is specifically designed for astrophysical settings with significant atmospheric stratification.
Central to the algorithm is a stratified background (or base) state density and pressure held in hydrostatic equilibrium, and also vary as a function of altitude and time.
MAESTRO is suitable for for spherical stars, as well as planar simulations of local dynamics.

@@

Since then, there have been many scientific investigations using MAESTRO, which include additional algorithmic enhancements.  Topics include:
\begin{itemize}
\item The convective phase preceding Chandrasekhar mass models for type Ia supernovae \citep{MAESTRO_convection,MAESTRO_AMR,MAESTRO_CASTRO}.
\item Convection in massive stars \citep{Gilet:2013}.
\item Sub-Chandrasekhar white dwarfs \citep{subChandra_I,subChandra_II}.  In the latter paper we introduced an optional modification to the momentum equation of velocity constraint that conserves total energy, can give results closer to compressible codes of convective dynamics, including low density regions at the surface of a star.
\item Type I X-ray bursts \citep{XRB_I,XRB_II,XRB_III}.  In \cite{XRB_I} we discuss the modifications required for implicit thermal conduction, as well as an additional ``volume discrepancy'' modification to the velocity constraint to force the species and enthalpy to evolve in a manner consistent with the background pressure.
\end{itemize}

In this paper, we characterize the performance of the MAESTROeX algorithm compared to the previous implementation.
We also develop simpler, alternative temporal integrators, with a longer term goal of developing high-order coupling schemes.
We also present a new treatment of the base state mapping to Cartesian grids for spherical problsm.

>>>>

Many astrophysical flows are highly subsonic; sound waves carry sufficiently little energy that they do not significantly affect the convective dynamics of the system.
In many of these flows, modeling long-time convective dynamics are of interest, and numerical approaches based on compressible hydrodynamics are intractable, even on modern supercomputers.
One approach to this problem is to use low Mach number models.
In a low Mach number approach, sound waves are eliminated from the governing equations while retaining compressibilitiy effects due to, e.g., nuclear energy release, compositional changes, and thermal diffusion.
The resulting model can be numerically integrated with much larger time steps than a compressible model.
Low Mach number models have been developed for a variety of contexts including combustion \citep{day2000numerical}, terrestrial atmospheric modeling \citep{duarte2015low}, 
and elastic solids \citep{abbate2017all}.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
The low Mach number model in MAESTRO is unique in that it is specifically designed for astrophysical settings with significant atmospheric stratification.
Central to the algorithm is a stratified background (or base) state density and pressure held in hydrostatic equilibrium, and also vary as a function of altitude and time.
MAESTRO is a structured-grid, finite-volume code that utlizes adaptive mesh refinement (AMR) to refine grids locally in space.
MAESTRO is suitable for for full spherical stars, as well as planar simulations of dynamics within localized regions of a star.

@@

Since then, there have been many scientific investigations using MAESTRO, which include additional algorithmic enhancements.  Topics include:
\begin{itemize}
\item The convective phase preceding Chandrasekhar mass models for type Ia supernovae \citep{MAESTRO_convection,MAESTRO_AMR,MAESTRO_CASTRO}.
\item Convection in massive stars \citep{Gilet:2013}.
\item Sub-Chandrasekhar white dwarfs \citep{subChandra_I,subChandra_II}.
%  In the latter paper we introduced an optional modification to the momentum equation of velocity constraint that conserves total energy, can give results closer to compressible codes of convective dynamics, including low density regions at the surface of a star.
\item Type I X-ray bursts \citep{XRB_I,XRB_II,XRB_III}.
%  In \cite{XRB_I} we discuss the modifications required for implicit thermal conduction, as well as an additional ``volume discrepancy'' modification to the velocity constraint to force the species and enthalpy to evolve in a manner consistent with the background pressure.
\end{itemize}

In this paper, we characterize the performance of the MAESTROeX algorithm compared to the previous implementation.
We also develop simpler, alternative temporal integrators, with a longer term goal of developing high-order coupling schemes.
We also present a new treatment of the base state mapping to Cartesian grids for spherical problems.

The resulting code is publicly available on GitHub at {\url https://github.com/AMReX-Astro/MAESTROeX} and uses the microphysics 
libraries at {\url https://github.com/starkiller-astro/Microphysics} and the AMReX software library for block-structured adaptive mesh refinement
{\url https://github.com/AMReX-Codes/amrex}.
